---
title: "Guided Factor Analysis on MCF10A Treated"
subtitle: "-- Genes Filtered by Detection Rate, NTC Regressed Out"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
wkdir <- "/project2/xinhe/yifan/Factor_analysis/GSE108699_MCF10A/"
```

```{r}
source(paste0(wkdir, "reports/analytics_functions.R"))
```

# Single cell expression data

CROP-seq data are from this paper: [On the design of CRISPR-based single cell molecular screens](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5882576/), GEO accession: GSE108699.

Perturbations:    
Knock-outs of 29 tumor-suppressor genes (TP53, ...), 1 non-targeting control.

Cells:    
MCF10A cells (normal human breast epithelial cells) with exposure to a DNA damaging agent, doxorubicin.   
Only cells with gRNA readout were kept, resulted in 5584 cells.

Genes:    
**Only genes detected in > 10% of cells were kept, resulted in 8046 genes.**

Normalization:    
Seurat "LogNormalize": log(count per 10K + 1).   
**Library size was regressed out, so were the non-targeting control (NTC) condition per cell. The residuals were used as input.**

```{r eval=FALSE, include=FALSE}
## Normalized expression:
scaled.gene_exp <- readRDS(paste0(wkdir, 'data/cropseq_dox.scaled_detect_01.corrected_lib_NTC.rds'))
```

```{r load meta data, fig.width=9}
metadata <- data.frame(fread(paste0(wkdir, 'data/GSE108699_cs.cell_metadata.tsv.gz'),
                             header = T, sep = '\t'), check.names = F)
dox_metadata <- metadata %>%
  filter(sample == "singles_dox_100nm_SI-3A-A6")
dox_metadata$cell_barcode <- sapply(strsplit(dox_metadata$cell, split = "_"),
                                    function(x){x[1]})
dox_metadata$guide_count[is.na(dox_metadata$guide_count)] <- 0
dox_metadata <- dox_metadata %>% filter(guide_count > 0)
```

# Factor correlation results

Guided sparse factor analysis, with the $G$ matrix being 30 types of KO conditions across cells.

```{r}
KO_names <- colnames(dox_metadata[, 17:46])
KO_names <- KO_names[KO_names != "NONTARGETING"]
all_KOs.G_mat <- as.matrix(dox_metadata[, KO_names] * 1)
```

## SVD Initialization
```{r}
All_KOs.gibbs_PM <- readRDS(paste0(wkdir, "gsfa_output_detect_01/no_NTC.corrected.gibbs_obj_k20.svd.PM.rds"))
All_KOs.gibbs_res_tb <- make_gibbs_res_tb(All_KOs.gibbs_PM, all_KOs.G_mat)
```

```{r fig.width=10}
heatmap_matrix <- All_KOs.gibbs_res_tb %>% select(starts_with("pval"))
rownames(heatmap_matrix) <- 1:nrow(heatmap_matrix)
colnames(heatmap_matrix) <- colnames(all_KOs.G_mat)
plot_pval_heatmap(heatmap_matrix,
                  factor_annot = All_KOs.gibbs_res_tb$pi)
no_NTC.heatmap_matrix <- heatmap_matrix
```

```{r fig.width=10, fig.height=5}
summ_pvalues(unlist(heatmap_matrix[, -29]), title_text = "GSFA All KOs (K=20, TP53 excluded)")
```

## Alternative Initializations and Chain Convergence

Other than initializing GSFA using truncated SVD, we also tried 2 runs with random initialization.

```{r}
rand_1.gibbs_PM <- 
  readRDS(paste0(wkdir, "gsfa_output_detect_01/no_NTC.corrected.gibbs_obj_k20.rand_01.PM.rds"))
rand_2.gibbs_PM <-
  readRDS(paste0(wkdir, "gsfa_output_detect_01/no_NTC.corrected.gibbs_obj_k20.rand_02.PM.rds"))
G0.gibbs_PM <-
  readRDS(paste0(wkdir, "gsfa_output_detect_01/G0.no_NTC.corrected.gibbs_obj_k20.svd.PM.rds"))
```

### Association results for one randomly initialized GSFA run

```{r  fig.width=10}
rand_1.gibbs_res_tb <- make_gibbs_res_tb(rand_1.gibbs_PM, all_KOs.G_mat)
rand_1.gibbs_res_tb <- rand_1.gibbs_res_tb %>% arrange(-sum_var)
heatmap_matrix <- rand_1.gibbs_res_tb %>% select(starts_with("pval"))
rownames(heatmap_matrix) <- rand_1.gibbs_res_tb$index
colnames(heatmap_matrix) <- colnames(all_KOs.G_mat)
plot_pval_heatmap(heatmap_matrix,
                  factor_annot = rand_1.gibbs_res_tb$pi)
```

```{r fig.width=10, fig.height=5}
summ_pvalues(unlist(heatmap_matrix[, -29]), title_text = "GSFA All KOs (K=20, TP53 excluded)")
```

### Chain mixing and convergence evaluation

To check the convergence of a certain scalar estimand., we adopt the potential scale reduction metric ($\hat{R}$) as described in Section 11.4 of _Bayesian Data Analysis, 3rd ed_.

The basic idea is to evaluate both the within-sequence variance and the between-sequence variance, and assess how close the two variances are.

If the potential scale reduction $\hat{R}$ is close to 1, it means our inference has reached a good level of convergence;   
if $\hat{R}$ is high, then we have reason to believe that proceeding with further simulations may improve our inference on this estimand.

Since the order and sign of factors are intercahngeable, we focus on evaluating the estimation of all pairs of $\beta_{m\cdot} \cdot W_{j\cdot}$ (a dot product that sums over all factors for each marker-gene pair).

We take the last 1000 stable samples from each of the 3 runs (1 SVD initialized, 2 randomly initialized). Each chain was further split into 2 halves, resulting in 6 chains with 500 iterations of samples each to assess convergence on.

The table below shows the summary of $\hat{R}$ of $\beta_{m\cdot} \cdot W_{j\cdot}$ across genes for a given KO condition $m$:

```{r show_R_hat_summary}
converg_mat <-
  readRDS(paste0(wkdir, "gsfa_output_detect_01/convergence/no_NTC.corrected.beta_W_converg_mat.1k-2k.rds"))
knitr::kable(signif(t(apply(converg_mat, 2, summary)), digits = 3)) %>%
  kable_styling() %>% scroll_box(width = '100%')
```

### Difference btw methods in factor estimation

For a pair of $Z$ estimations from 2 inference methods, $Z_1, Z_2$, we quantify the pairwise estimation difference as $||Z_1Z_1^T - Z_2Z_2^T||_1/N^2$, where $N$ is the number of rows (samples) in $Z$.

```{r}
cat("Guided SVD vs Rand_01:",
    ZZT_mean_diff(All_KOs.gibbs_PM$Z.pm, rand_1.gibbs_PM$Z.pm) %>% signif(digits = 3))
cat("\n\n")
cat("Guided SVD vs Rand_02:",
    ZZT_mean_diff(All_KOs.gibbs_PM$Z.pm, rand_2.gibbs_PM$Z.pm) %>% signif(digits = 3))
cat("\n\n")
cat("Rand_01 vs Rand_02:",
    ZZT_mean_diff(rand_1.gibbs_PM$Z.pm, rand_2.gibbs_PM$Z.pm) %>% signif(digits = 3))
cat("\n\n")
cat("Guided SVD vs Unguided:",
    ZZT_mean_diff(All_KOs.gibbs_PM$Z.pm, G0.gibbs_PM$Z.pm) %>% signif(digits = 3))
cat("\n\n")
```

```{r G0 heatmap}
G0.gibbs_res_tb <- make_gibbs_res_tb(G0.gibbs_PM, all_KOs.G_mat)
heatmap_matrix <- G0.gibbs_res_tb %>% select(starts_with("pval"))
rownames(heatmap_matrix) <- 1:nrow(heatmap_matrix)
colnames(heatmap_matrix) <- colnames(all_KOs.G_mat)
plot_pval_heatmap(heatmap_matrix,
                  factor_annot = G0.gibbs_res_tb$pi)
G0.heatmap_matrix <- heatmap_matrix
```

## Comparison with NTC included GSFA result

```{r}
NTC_included.gibbs_PM <-
  readRDS(paste0(wkdir, "gsfa_output_detect_01/All.corrected.gibbs_obj_k20.svd.restart_01.PM.rds"))
NTC_included.gibbs_res_tb <- make_gibbs_res_tb(NTC_included.gibbs_PM, all_KOs.G_mat)
NTC_included.heatmap <- NTC_included.gibbs_res_tb %>% select(starts_with("pval"))
rownames(NTC_included.heatmap) <- 1:nrow(NTC_included.heatmap)
colnames(NTC_included.heatmap) <- colnames(all_KOs.G_mat)
```

```{r pval_comparison, eval=FALSE, include=FALSE}
plot_lst <- list()
for (i in 1:ncol(heatmap_matrix)){
  pval_df <- data.frame(orig = -log10(NTC_included.heatmap[, i]),
                        no_NTC = -log10(heatmap_matrix[, i]))
  p <- ggplot(pval_df, aes_string(x = "orig", y = "no_NTC")) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(title = colnames(heatmap_matrix)[i]) +
    theme(axis.title = element_blank())
  plot_lst[[i]] <- p
}
args <- c(plot_lst, list(nrow = 5,
                         left = "New (NTC regressed)", bottom = "Original",
                         top = "Factor ~ Marker, -log10(P value)"))
do.call(grid.arrange, args)
```

```{r fig.width=12}
pval_vec_1 <- unlist(no_NTC.heatmap_matrix)
name_1 <- "NTC regressed"
pval_vec_2 <- unlist(NTC_included.heatmap)
name_2 <- "NTC included"
paired_df <- data.frame(V1 = pval_vec_1[order(pval_vec_1)],
                        V2 = pval_vec_2[order(pval_vec_2)])
paired_df <- reshape2::melt(paired_df, value.name = "pval")
paired_df <- paired_df %>%
  mutate(neg_logp = -log10(pval)) %>%
  mutate(rank = rep(1:length(pval_vec_1), 2)) %>%
  rename(type = variable) %>%
  mutate(type = factor(type)) %>%
  mutate(type = recode(type, V1 = name_1, V2 = name_2))
p1 <- ggplot(paired_df, aes(x = rank, y = neg_logp, color = factor(type))) +
  geom_point(size = 0.8) +
  labs(y = "-log10(P value)") +
  theme(legend.title = element_blank(),
        legend.position = "none")
p2 <- ggplot(paired_df, aes(x = rank, y = neg_logp, color = type)) +
  geom_point(size = 0.8) +
  scale_y_continuous(limits = c(0, 20)) +
  labs(y = "-log10(P value)") +
  theme(legend.title = element_blank())
grid.arrange(p1, p2, nrow = 1)
```

# Gene Ontology Enrichment Analysis

Genes w/ non-zero loadings vs all genes selected for factor analysis.

GO category: Biological Process.

## PIP cutoff at 0.5

GO terms that passed overrpresentation analysis fold change $\geq$ 2 and q value $<$ 0.05:

```{r eval=FALSE, include=FALSE}
gsfa.go_enrich <-
  readRDS(paste0(wkdir, "gsfa_output_detect_01/enrich_res/All_KOs.corrected.k20_svd.go_enrich.cutoff_0.5.rds"))
signif_num <- print_enrich_tb(gsfa.go_enrich)
signif_num.df <- t(data.frame(Factor = 1:length(signif_num),
                              Signif_GO_terms = signif_num,
                              row.names = NULL))
knitr::kable(rbind(signif_num.df[, 1:10],
                   signif_num.df[, 11:20])) %>%
  kable_styling() %>% scroll_box(width = '100%')
```
