---
title: "Permutation Analysis on Factors Obtain from MCF10A Treated Expression Data"
subtitle: "-- P value ~ KO Associations"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
wkdir <- "/project2/xinhe/yifan/Factor_analysis/GSE108699_MCF10A/"
```

```{r}
source(paste0(wkdir, "reports/analytics_functions.R"))
```

For each factor analysis result, we did 1000 permutaions on the cell labels, and correlated the permuted KO conditions with the factors.

The empirical p value distribution is plotted by combining the association p values between the 20 factors and the 30 knock-out (KO) conditions over 1000 permutations ($20\times 30\times 1000 = 6\times 10^5$ p values in total).

Conclusion:

* For factors obtained by GSFA/FLASH on **genes filtered by detection rate**, they do not have inflated associations with KO conditions.   
* For factors obtained by GSFA/FLASH on **genes filtered by variance**, the later factors (sparser and have smaller PVEs) have uncalibrated associations with KO conditions, with inflation both near 0 and toward 1.

# Genes Filtered by Detection Rate

**As shown below, permutated p value distribution is generally uniform for all factors.**

## GSFA Detection > 10% (8046 genes)

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "gsfa_output_detect_01/",
                            "permutation/All.gibbs_k20_svd.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "GSFA Detection > 10% Genes\n(20 Factors, 30 KOs)")
```

P value distribution by each factor:

```{r fig.width=12, fig.height=10}
plot_pval_list_grid(pval_list)
```

## FLASH Detection > 10%

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "flash_output/permutation/",
                            "detect_0.1.All_flash_k20.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "FLASH Detection > 10% Genes\n(20 Factors, 30 KOs)")
```

## GSFA Detection > 5% (9895 genes)

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "gsfa_output_detect_005/",
                            "permutation/All.gibbs_k20_svd.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "GSFA Detection > 5% Genes\n(20 Factors, 30 KOs)")
```

## FLASH Detection > 5%

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "flash_output/permutation/",
                            "detect_0.05.All_flash_k20.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "FLASH Detection > 5% Genes\n(20 Factors, 30 KOs)")
```

# Genes Filtered by Variance

**As shown below, permutated p value distribution deviates greatly from uniform for some later factors.**

## GSFA on Top 8K Variable Genes

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "gsfa_output_top8k_var/",
                            "permutation/All.gibbs_k20_svd.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "GSFA Top 8K Variable Genes\n(20 Factors, 30 KOs)")
```

P value distribution by each factor:

```{r fig.width=12, fig.height=10}
plot_pval_list_grid(pval_list)
```

## FLASH on Top 8K Variable Genes

```{r fig.height=5, fig.width=10}
pval_list <- readRDS(paste0(wkdir, "flash_output/permutation/",
                            "top8k_var.All_flash_k20.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "FLASH Top 8K Variable Genes\n(20 Factors, 30 KOs)")
```

P value distribution by each factor:

```{r fig.width=12, fig.height=10}
plot_pval_list_grid(pval_list)
```

## GSFA on Top 10K Variable Genes

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "gsfa_output_top10k_var/",
                            "permutation/All.gibbs_k20_svd.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "GSFA Top 10K Variable Genes\n(20 Factors, 30 KOs)")
```

P value distribution by each factor:

```{r fig.width=12, fig.height=10}
plot_pval_list_grid(pval_list)
```


## FLASH on Top 10K Variable Genes

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "flash_output/permutation/",
                            "top10k_var.All_flash_k20.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "FLASH Top 10K Variable Genes\n(20 Factors, 30 KOs)")
```

P value distribution by each factor:

```{r fig.width=12, fig.height=10}
plot_pval_list_grid(pval_list)
```

# SCTransform Normalization

Instead of logCPM normalization on the raw count matrix, we also tried sctransform on the data before our factor analysis.

However, the inflation in factor ~ KO association seems to be worse.

## GSFA Detection > 10% (8046 genes)

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "gsfa_output_sctran_detect_01/",
                            "permutation/All.gibbs_k20_svd.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "GSFA Detection > 10% Genes\n(20 Factors, 30 KOs)")
```

## GSFA on Top 8K Variable Genes

```{r fig.width=10, fig.height=5}
pval_list <- readRDS(paste0(wkdir, "gsfa_output_sctran_top8k_var/",
                            "permutation/All.gibbs_k20_svd.pval_list.rds"))
summ_pvalues(unlist(pval_list), title_text = "GSFA Top 8K Variable Genes\n(20 Factors, 30 KOs)")
```
