---
title: "Unguided Factor Analysis on MCF10A Treated"
subtitle: "-- Genes Filtered by Detection Rate"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
wkdir <- "/project2/xinhe/yifan/Factor_analysis/GSE108699_MCF10A/"
```

```{r}
source(paste0(wkdir, "reports/analytics_functions.R"))
```

# Single cell expression data

CROP-seq data are from this paper: [On the design of CRISPR-based single cell molecular screens](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5882576/), GEO accession: GSE108699.

Perturbations:    
Knock-outs of 29 tumor-suppressor genes (TP53, ...), 1 non-targeting control.

Cells:    
MCF10A cells (normal human breast epithelial cells) with exposure to a DNA damaging agent, doxorubicin.   
Only cells with gRNA readout were kept, resulted in 5584 cells.

Genes:    
**Only genes detected in > 10% of cells were kept, resulted in 8046 genes.**

Normalization:    
Seurat "LogNormalize": log(count per 10K + 1).   
**Library size was regressed out, and the residuals were used as input.**

```{r eval=FALSE, include=FALSE}
## Normalized expression:
scaled.gene_exp <- readRDS(paste0(wkdir, 'data/cropseq_dox.scaled_genes_0.1.corrected_lib_size.rds'))
```

```{r load meta data, fig.width=9}
metadata <- data.frame(fread(paste0(wkdir, 'data/GSE108699_cs.cell_metadata.tsv.gz'),
                             header = T, sep = '\t'), check.names = F)
dox_metadata <- metadata %>%
  filter(sample == "singles_dox_100nm_SI-3A-A6")
dox_metadata$cell_barcode <- sapply(strsplit(dox_metadata$cell, split = "_"),
                                    function(x){x[1]})
dox_metadata$guide_count[is.na(dox_metadata$guide_count)] <- 0
dox_metadata <- dox_metadata %>% filter(guide_count > 0)
```

# Factor correlation results

Unguided sparse factor analysis, with $G$ being 0s.

```{r}
all_KOs.G_mat <- as.matrix(dox_metadata[, 17:46] * 1)
```

## SVD Initialization

```{r}
G0.gibbs_PM <-  readRDS(paste0(wkdir, "gsfa_output_detect_01/G0.corrected.gibbs_obj_k20.svd.PM.rds"))
G0.gibbs_res_tb <- make_gibbs_res_tb(G0.gibbs_PM, all_KOs.G_mat)
```

```{r fig.width=10}
heatmap_matrix <- G0.gibbs_res_tb %>% select(starts_with("pval"))
rownames(heatmap_matrix) <- 1:nrow(heatmap_matrix)
colnames(heatmap_matrix) <- colnames(all_KOs.G_mat)
ntc_inx <- which(colnames(heatmap_matrix) == "NONTARGETING")
heatmap_matrix <- heatmap_matrix[, c(colnames(heatmap_matrix)[ntc_inx],
                                     colnames(heatmap_matrix)[-ntc_inx])]
plot_pval_heatmap(heatmap_matrix,
                  factor_annot = G0.gibbs_res_tb$pi)
```

```{r fig.width=10, fig.height=5}
summ_pvalues(unlist(heatmap_matrix[, -30]), title_text = "Unguided All KOs (K=20, TP53 excluded)")
```

# Gene Ontology Enrichment Analysis

Genes w/ non-zero loadings vs all genes selected for factor analysis.

GO category: Biological Process.

## PIP cutoff at 0.5

GO terms that passed overrpresentation analysis fold change $\geq$ 2 and q value $<$ 0.05:

```{r}
G0.go_enrich <-
  readRDS(paste0(wkdir, "gsfa_output_detect_01/enrich_res/G0.corrected.k20_svd.go_enrich.cutoff_0.5.rds"))
signif_num <- print_enrich_tb(G0.go_enrich)
signif_num.df <- t(data.frame(Factor = 1:length(signif_num),
                              Signif_GO_terms = signif_num,
                              row.names = NULL))
knitr::kable(rbind(signif_num.df[, 1:10],
                   signif_num.df[, 11:20])) %>%
  kable_styling() %>% scroll_box(width = '100%')
```
