---
title: "MUSIC Topic Modeling of LUHMES CROP-seq Data"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 13),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)
library(clusterProfiler)
# library(MUSIC)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/analytics_functions.R")
wkdir <- "/project2/xinhe/yifan/Factor_analysis/LUHMES/"
gsfa_folder <- "gsfa_output_detect_01/merged_new/"
guide <- "All"
pip_col_fun <- circlize::colorRamp2(breaks = c(0, 0.5, 1),
                                    colors = c("black", "purple", "gold"))
```

## Single cell expression data

Source:   
[High-throughput single-cell functional elucidation of neurodevelopmental disease-associated genes reveals convergent mechanisms altering neuronal differentiation](https://genome.cshlp.org/content/30/9/1317.full), GEO accession: GSE142078.

Perturbations:    
CRISPR knock-down of 14 autism spectrum disorder (ASD)–associated genes (3 gRNAs per gene) + 5 non-targeting gRNAs.

Cells:    
Lund human mesencephalic (LUHMES) neural progenitor cell line.   
Cells from 3 batches were merged together into 1 analysis. All cells have only a single type of gRNA readout.

## MUSIC pipeline

### Preprocessing

* Cell quality control    
* SAVER data imputation   
* Filtering cells with invalid edits    
(if a cell's differentially expressed gene profile is more similar to the control
cells than to the other cells with the same perturbation.)  
* Selecting genes with large dispersion difference between case and control  
* Normalizing and rounding the expression values to non-negative integers

Final gene and cell size:
```{r}
crop_seq_vargene <- readRDS(paste0(wkdir, "music_output/music_vargene.merged.rds"))
print(dim(crop_seq_vargene$expression_profile))
perturb_information <- crop_seq_vargene$perturb_information
```

### Result interpretation

**Annotating the functions of each topic:**

MUSIC obtains the occurrence probabilities of genes available in each topic. It then selects the top 10% genes of each topic based on their occurrence probabilities, and perform functional enrichment analysis using all genes in topic modeling as background.

**Characterizing topc-perturbation relationships:**

For a specific topic, MUSIC prioritizes the perturbation effect by calculating the specific topic probability difference (TPD) between the case and control groups.

For the $i$-th perturbation on the $j$-th topic, the TPD against the control group is computed as the Student's $t$-statistics between $\{z_{mj}\}_{m\in \text{perturbtion }i}$ and $\{z_{nj}\}_{n\in \text{control}}$,  
where $z_{mj}$ is the probability of topic $j$ in cell $m$ normalized _w.r.t._ the control group: $z_{mj} = \frac{\theta_{mj}-\mu_{\text{control}}}{\sigma_{\text{control}}}$.

```{r}
guide_mat <- data.frame(matrix(nrow = length(perturb_information),
                               ncol = length(names(table(perturb_information)))))
rownames(guide_mat) <- names(perturb_information)
colnames(guide_mat) <- names(table(perturb_information))
for (i in colnames(guide_mat)){
  guide_mat[[i]] <- (perturb_information == i) * 1
}
```

## 5 Topics

```{r}
topic_res <- readRDS(paste0(wkdir, "music_output/music_merged_5_topics.rds"))
topic_res_annot <- readRDS(paste0(wkdir, "music_output/enrich_res/",
                                 "5_topics.go_enrich.top_10_percent.rds"))
```

### Topic annotations

```{r fig.height=6, fig.width=10}
signif_res <- plot_top_enrich_terms(topic_res_annot, top_num = 5, plot = T)
```

```{r}
signif_num.df <- t(data.frame(Topic = 1:length(signif_res$signif_num),
                              Signif_GO_terms = signif_res$signif_num,
                              row.names = NULL))
knitr::kable(signif_num.df,
             caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '60%')
```

**Neural-related topic(s):**

Topic 1

### Topic-perturbation relationship

```{r fig.width=6, fig.height=5}
distri_diff <- MUSIC::Diff_topic_distri(topic_res$models[[1]],
                                        perturb_information, plot = F)
t_D_diff_matrix <- dcast(distri_diff %>% dplyr::select(knockout, variable, t_D_diff),
                         knockout ~ variable)
rownames(t_D_diff_matrix) <- t_D_diff_matrix$knockout
t_D_diff_matrix$knockout <- NULL
Heatmap(t_D_diff_matrix,
        name = "Topic probability difference\n(vs ctrl)",
        cluster_rows = T, cluster_columns = T,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
```

```{r fig.width=5, fig.height=5}
association_list <- factor_matrix_regression(topic_res$models[[1]]@gamma, guide_mat)
pval_mat <- association_list$pval
rownames(pval_mat) <- paste("Topic", 1:nrow(pval_mat))
colnames(pval_mat) <- colnames(guide_mat)
Heatmap(-log10(t(pval_mat)),
        name = "-log10(Topic~Guide\np value)",
        col = circlize::colorRamp2(c(0, 3, 15), c("blue3", "white", "firebrick")),
        row_title = "KD Perturbations", column_title = "Topic",
        cluster_rows = F, cluster_columns = F,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
```

## 10 Topics

```{r}
topic_res <- readRDS(paste0(wkdir, "music_output/music_merged_10_topics.rds"))
topic_res_annot <- readRDS(paste0(wkdir, "music_output/enrich_res/",
                                 "10_topics.go_enrich.top_10_percent.rds"))
```

### Topic annotations

```{r fig.height=9, fig.width=10}
signif_res <- plot_top_enrich_terms(topic_res_annot, top_num = 5, plot = T)
```

```{r}
signif_num.df <- t(data.frame(Topic = 1:length(signif_res$signif_num),
                              Signif_GO_terms = signif_res$signif_num,
                              row.names = NULL))
knitr::kable(signif_num.df[, 1:10],
             caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '100%')
```

**Neural-related topic(s):**

Topic 4, 5, 10

### Topic-perturbation relationship

```{r fig.width=8, fig.height=5}
distri_diff <- MUSIC::Diff_topic_distri(topic_res$models[[1]],
                                        perturb_information, plot = F)
t_D_diff_matrix <- dcast(distri_diff %>% dplyr::select(knockout, variable, t_D_diff),
                         knockout ~ variable)
rownames(t_D_diff_matrix) <- t_D_diff_matrix$knockout
t_D_diff_matrix$knockout <- NULL
Heatmap(t_D_diff_matrix,
        name = "Topic probability difference\n(vs ctrl)",
        cluster_rows = T, cluster_columns = T,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
```

```{r fig.width=7, fig.height=5}
association_list <- factor_matrix_regression(topic_res$models[[1]]@gamma, guide_mat)
pval_mat <- association_list$pval
rownames(pval_mat) <- paste("Topic", 1:nrow(pval_mat))
colnames(pval_mat) <- colnames(guide_mat)
Heatmap(-log10(t(pval_mat)),
        name = "-log10(Topic~Guide\np value)",
        col = circlize::colorRamp2(c(0, 3, 15), c("blue3", "white", "firebrick")),
        row_title = "KD Perturbations", column_title = "Topic",
        cluster_rows = F, cluster_columns = F,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
```

## 20 Topics

```{r}
topic_res <- readRDS(paste0(wkdir, "music_output/music_merged_20_topics.rds"))
topic_res_annot <- readRDS(paste0(wkdir, "music_output/enrich_res/",
                                 "20_topics.go_enrich.top_10_percent.rds"))
```

### Topic annotations

```{r fig.height=12, fig.width=12}
signif_res <- plot_top_enrich_terms(topic_res_annot, top_num = 5, plot = T)
```

```{r}
signif_num.df <- t(data.frame(Topic = 1:length(signif_res$signif_num),
                              Signif_GO_terms = signif_res$signif_num,
                              row.names = NULL))
knitr::kable(rbind(signif_num.df[, 1:10], signif_num.df[, 11:20]),
             caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '100%')
```

**Neural-related topic(s):**

Topic 1, 4, 9

### Topic-perturbation relationship

```{r fig.width=12, fig.height=5}
distri_diff <- MUSIC::Diff_topic_distri(topic_res$models[[1]],
                                        perturb_information, plot = F)
t_D_diff_matrix <- dcast(distri_diff %>% dplyr::select(knockout, variable, t_D_diff),
                         knockout ~ variable)
rownames(t_D_diff_matrix) <- t_D_diff_matrix$knockout
t_D_diff_matrix$knockout <- NULL
Heatmap(t_D_diff_matrix,
        name = "Topic probability difference\n(vs ctrl)",
        cluster_rows = T, cluster_columns = T,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
```

```{r fig.width=12, fig.height=5}
association_list <- factor_matrix_regression(topic_res$models[[1]]@gamma, guide_mat)
pval_mat <- association_list$pval
rownames(pval_mat) <- paste("Topic", 1:nrow(pval_mat))
colnames(pval_mat) <- colnames(guide_mat)
Heatmap(-log10(t(pval_mat)),
        name = "-log10(Topic~Guide\np value)",
        col = circlize::colorRamp2(c(0, 3, 15), c("blue3", "white", "firebrick")),
        row_title = "KD Perturbations", column_title = "Topic",
        cluster_rows = F, cluster_columns = F,
        column_names_rot = 45,
        heatmap_legend_param = list(title_gp = gpar(fontsize = 12, fontface = "bold")))
```

