---
title: "Functional Analysis on LUHMES CROP-seq GSFA Results"
subtitle: "-- NTC included, unique UMI count corrected"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/analytics_functions.R")
print_signif_tb <- function(signif_num, caption_text){
  if (length(signif_num) < 15){
    signif_num <- c(signif_num, rep(NA, 15 - length(signif_num)))
  }
  signif_num <- t(data.frame(KO = names(signif_num),
                             Num_genes = signif_num,
                             row.names = NULL))
  print(knitr::kable(rbind(signif_num[, 1:5], signif_num[, 6:10], signif_num[, 11:15]),
                     caption = caption_text) %>%
    kable_styling() %>% scroll_box(width = '100%'))
}
print_lfsr_signif_tb <- function(lfsr_mat, lfsr_cutoff, caption_text){
  signif_num <- colSums(lfsr_mat < lfsr_cutoff)
  print_signif_tb(signif_num, caption_text)
}

wkdir <- "/project2/xinhe/yifan/Factor_analysis/LUHMES/"
gsfa_folder <- "gsfa_output_detect_01/merged_new/"
guide <- "All"
pip_col_fun <- circlize::colorRamp2(breaks = c(0, 0.5, 1),
                                    colors = c("black", "purple", "gold"))
```

# GSFA on LUHMES CROP-seq Data

## Data Processing

Cells:    
LUHMES cells from 3 batches were merged together into 1 analysis. All cells have only a single type of gRNA readout. Quality control resulted in 8708 cells.

Genes:    
**Only genes detected in > 10% of cells were kept, resulted in 6213 genes.**

Normalization:    
Seurat "LogNormalize": log(count per 10K + 1).   
**Batch effect, unique UMI count, library size, and mitochondria percentage were all corrected for. The corrected and scaled expression data were used as input for subsequent factor analysis.**

## GSFA Results (SVD Initialization)

Here, our "guide", $G$ matrix, consists of 15 types (14 genes + NTC) of gene-level knock-down conditions across cells.

```{r load meta data}
metadata <- readRDS(paste0(wkdir, "processed_data/merged_metadata.rds"))
all_KOs.G_mat <- metadata[, 4:18]
KO_names <- colnames(all_KOs.G_mat)
if (guide == "no_NTC"){
  KO_names <- KO_names[KO_names != "Nontargeting"]
  all_KOs.G_mat <- all_KOs.G_mat[, KO_names]
}
nmarkers <- ncol(all_KOs.G_mat)
```

Gibbs sampling was initiated from SVD, and conducted for 2000 iterations; the posterior mean estimates were averaged over the last 500 iterations.

```{r svd_tb}
All_KOs.gibbs_PM <- readRDS(paste0(wkdir, gsfa_folder, guide, ".gibbs_obj_k20.svd.restart.PM.rds"))
stopifnot(nrow(All_KOs.gibbs_PM$beta_pm) == nmarkers + 1)
All_KOs.gibbs_res_tb <- make_gibbs_res_tb(All_KOs.gibbs_PM, all_KOs.G_mat, compute_pve = F)
```

### Factor ~ KO Beta PIP
```{r fig.width=8}
beta_pip_matrix <- t(All_KOs.gibbs_PM$Gamma_pm[-(nmarkers + 1), ])
rownames(beta_pip_matrix) <- 1:nrow(beta_pip_matrix)
colnames(beta_pip_matrix) <- colnames(all_KOs.G_mat)
Heatmap(beta_pip_matrix, name = "Beta PIP", col = pip_col_fun,
        row_title = "Factors",  column_title = "Gene Perturbations",
        cluster_rows = F, cluster_columns = F)
```

### DEGs (LFSR < 0.05) under Each Perturbation

```{r}
lfsr_cutoff <- 0.05
lfsr_mat <- readRDS(paste0(wkdir, gsfa_folder, guide, ".gibbs_obj_k20.svd.restart.lfsr_mat.rds"))
print_lfsr_signif_tb(lfsr_mat, lfsr_cutoff,
                     paste0("GSFA, # of genes that passed LFSR cutoff of 0.05:"))
```

Overlap of DEGs between perturbations:

```{r fig.width=7, fig.height=5.5}
num_signif_genes <- colSums(lfsr_mat < lfsr_cutoff)
signif_markers <- names(num_signif_genes)[which(num_signif_genes > 0)]
plot_pairwise.corr_heatmap(lfsr_mat[, signif_markers] < lfsr_cutoff, corr_type = "jaccard",
                           label_size = 11,
                           color_vec = c("#FFEDE6", "#FFA584", "#FF7F50"))
```

Result of knock-down targets:

```{r knockdown_targets}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/gene_discovery_functions.R")
feature.names <- data.frame(fread(paste0(wkdir, "GSE142078_raw/GSM4219576_Run2_genes.tsv.gz"),
                                  header = FALSE), stringsAsFactors = FALSE)
names(feature.names) <- c("ID", "Name")

gsfa.DE_list <- compute_beta_dot_W(lfsr_mat, All_KOs.gibbs_PM)
gsfa.DE_list <- lapply(gsfa.DE_list, function(x){ left_join(x, feature.names,
                                                            by = c("gene_ID" = "ID")) })
for (m in signif_markers){
  if (!(m %in% gsfa.DE_list[[m]]$Name)){
    cat(paste0(m, " is not detected in > 10% cells."))
    cat("\n")
  } else {
    print(kable(gsfa.DE_list[[m]] %>% filter(Name == m), digits = 3,
              caption = paste0(m, " knock-down")) %>%
    kable_styling())
  }
}
```

# Gene Set Enrichment Analysis

## Enrichment by factor

Target genes: Genes w/ non-zero loadings in each factor (PIP cutoff at 0.95);    
Backgroud genes: all 6213 genes included in factor analysis;    
Statistical test: hypergeometric test (over-representation test);    
Only GO terms/pathways within the size of 10 ~ 500 and have a test FDR $<$ 0.05 are shown in the webpages.

### Reactome Pathway Over-Representation Analysis

Gene sets: The [Reactome pathway database](http://www.reactome.org/).   

**Factors of interest**

* [Factor 8](http://www.webgestalt.org/results/1614728526/#): p53-Independent DNA Damage Response, G2/M Transition
* [Factor 9](http://www.webgestalt.org/results/1614728715/#): Neuronal System, Transmission across Chemical Synapses
* [Factor 12](http://www.webgestalt.org/results/1614728793/#): Gap junction trafficking
* [Factor 13](http://www.webgestalt.org/results/1614728429/#): Neurotransmitter receptors and postsynaptic signal transmission, Ion channel transport, Neuronal System
* [Factor 15](http://www.webgestalt.org/results/1614728922/#): Transmission across Chemical Synapses, Ca2+ pathway, Gap junction trafficking
* [Factor 17](http://www.webgestalt.org/results/1614729171/#): Axon guidance
* [Factor 19](http://www.webgestalt.org/results/1614729226/#): Activation of NMDA receptors and postsynaptic events, Mitotic phases, Cell Cycle Checkpoints
* [Factor 20](http://www.webgestalt.org/results/1614729303/#): Neurotransmitter receptors and postsynaptic signal transmission, Ion channel transport

### GO Slim Over-Representation Analysis

Gene sets: [Gene ontology](http://www.geneontology.org/) "Biological Process" (non-redundant).

**Factors of interest**

* [Factor 10](http://www.webgestalt.org/results/1614731260/#): nerve development, cell fate commitment;   
* [Factor 11](http://www.webgestalt.org/results/1614731201/#): chromatin remodeling, DNA recombination;   
* [**Factor 13**](http://www.webgestalt.org/results/1614731171/#): regulation of metal ion transport, axon development, ...;   
* [**Factor 15**](http://www.webgestalt.org/results/1614731117/#): neuron migration, response to metal ion, CNS neuron development, ...;   
* [**Factor 18**](http://www.webgestalt.org/results/1614730861/#): CNS neuron differentiation, xxx tissue development, ...;   
* [Factor 19](http://www.webgestalt.org/results/1614730760/#): axon development;   
* [**Factor 20**](http://www.webgestalt.org/results/1614730397/#): nerve development, neurotransmitter transport, ....

## Enrichment by perturbation

Target genes: Genes w/ LSFR < 0.05 under each perturbation;    
Backgroud genes: all 6213 genes included in factor analysis;    
Statistical test: hypergeometric test (over-representation test);    
Only GO terms/pathways within the size of 10 ~ 500 and have a test FDR $<$ 0.05 are shown in the webpages.

### Reactome Pathway Over-Representation Analysis

* [ADNP](http://www.webgestalt.org/results/1614735983/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614748228/#)
    * Down-regulated
* [ARID1B](http://www.webgestalt.org/results/1614736635/#)
    * Up-regulated
    * [Down-regulated](http://www.webgestalt.org/results/1614748345/#)
* [ASH1L](http://www.webgestalt.org/results/1614736695/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614748437/#)
    * [Down-regulated](http://www.webgestalt.org/results/1614748499/#): mRNA Splicing, Transcriptional Regulation by TP53
* [CHD2](http://www.webgestalt.org/results/1614737079/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614748559/#)
    * [Down-regulated](http://www.webgestalt.org/results/1614748857/#): Axon guidance
* [PTEN](http://www.webgestalt.org/results/1614737261/#)
    * Up-regulated
    * [Down-regulated](http://www.webgestalt.org/results/1614749222/#): Axon guidance
* [SETD5](http://www.webgestalt.org/results/1614737506/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614749447/#)
    * [Down-regulated](http://www.webgestalt.org/results/1614749474/#): Axon guidance

### GO Slim Over-Representation Analysis

* [ADNP](http://www.webgestalt.org/results/1614737785/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614741328/#):regulation of metal ion transport, regulation of neurotransmitter levels
    * Down-regulated
* [ARID1B](http://www.webgestalt.org/results/1614737823/#)
    * Up-regulated
    * [Down-regulated](http://www.webgestalt.org/results/1614741644/#): regulation of metal ion transport
* [ASH1L](http://www.webgestalt.org/results/1614737877/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614741821/#): regulation of metal ion transport
    * [Down-regulated](http://www.webgestalt.org/results/1614741971/#)
* [CHD2](http://www.webgestalt.org/results/1614737909/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614742122/#)
    * [Down-regulated](http://www.webgestalt.org/results/1614742235/#): central nervous system neuron differentiation
* [PTEN](http://www.webgestalt.org/results/1614737973/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614742400/#): regulation of ion transmembrane transport, regulation of neuron projection development
    * [Down-regulated](http://www.webgestalt.org/results/1614743125/#): regulation of metal ion transport
* [SETD5](http://www.webgestalt.org/results/1614738030/#)
    * [Up-regulated](http://www.webgestalt.org/results/1614743225/#): axon development, positive regulation of neurogenesis
    * [Down-regulated](http://www.webgestalt.org/results/1614743342/#)

    