---
title: "Guided Factor Analysis on LUHMES CROP-seq Data"
subtitle: "-- Permutation (NTC included, unique UMI count corrected)"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)
library(clusterProfiler)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/analytics_functions.R")
print_signif_tb <- function(signif_num, caption_text){
  if (length(signif_num) < 15){
    signif_num <- c(signif_num, rep(NA, 15 - length(signif_num)))
  }
  signif_num <- t(data.frame(KO = names(signif_num),
                             Num_genes = signif_num,
                             row.names = NULL))
  print(knitr::kable(rbind(signif_num[, 1:5], signif_num[, 6:10], signif_num[, 11:15]),
                     caption = caption_text) %>%
    kable_styling() %>% scroll_box(width = '100%'))
}
print_lfsr_signif_tb <- function(lfsr_mat, lfsr_cutoff, caption_text){
  signif_num <- colSums(lfsr_mat < lfsr_cutoff)
  print_signif_tb(signif_num, caption_text)
}

wkdir <- "/project2/xinhe/yifan/Factor_analysis/LUHMES/"
gsfa_folder <- "gsfa_output_detect_01/merged_new/"
guide <- "All"
pip_col_fun <- circlize::colorRamp2(breaks = c(0, 0.5, 1),
                                    colors = c("black", "purple", "gold"))
```

# Single cell expression data

Source:   
[High-throughput single-cell functional elucidation of neurodevelopmental disease-associated genes reveals convergent mechanisms altering neuronal differentiation](https://genome.cshlp.org/content/30/9/1317.full), GEO accession: GSE142078.

Perturbations:    
CRISPR knock-down of 14 autism spectrum disorder (ASD)–associated genes (3 gRNAs per gene) + 5 non-targeting gRNAs.

Cells:    
Lund human mesencephalic (LUHMES) neural progenitor cell line.   
Cells from 3 batches were merged together into 1 analysis. All cells have only a single type of gRNA readout. Quality control resulted in 8708 cells.

Genes:    
**Only genes detected in > 10% of cells were kept, resulted in 6213 genes.**

Normalization:    
Seurat "LogNormalize": log(count per 10K + 1).   
**Batch effect, unique UMI count, library size, and mitochondria percentage were all corrected for. The corrected and scaled expression data were used as input for subsequent factor analysis.**

```{r load meta data, fig.width=9}
metadata <- readRDS(paste0(wkdir, "processed_data/merged_metadata.rds"))
all_KOs.G_mat <- metadata[, 4:18]
KO_names <- colnames(all_KOs.G_mat)
nmarkers <- ncol(all_KOs.G_mat)
```

# Permutation

Cell labels in the expression data were permuted randomly so that they are no longer correlated with the knock-down conditions. Then GSFA was performed still using all conditions as guides. Factor-guide association as well as the LFSR of each gene were evaluated as usual.

In total, 10 random permutation rounds like this were conducted.

```{r load PM results}
gibbs_lst <- list()
for (i in 1:10){
  perm_PM <- readRDS(paste0(wkdir, gsfa_folder, "All.gibbs_obj_k20.perm_", i, ".PM.rds"))
  gibbs_lst[[i]] <- perm_PM
}
```

## Factor ~ KO Beta PIP
```{r fig.width=8}
for (i in 1:length(gibbs_lst)){
  perm_PM <- gibbs_lst[[i]]
  beta_pip_matrix <- t(perm_PM$Gamma_pm[-(nmarkers + 1), ])
  rownames(beta_pip_matrix) <- 1:nrow(beta_pip_matrix)
  colnames(beta_pip_matrix) <- colnames(all_KOs.G_mat)
  hm <- Heatmap(beta_pip_matrix, name = "Beta PIP", col = pip_col_fun,
                row_title = "Factors",  column_title = "KO Perturbations",
                cluster_rows = F, cluster_columns = F)
  draw(hm)
}
```

## Factor ~ KO Posterior Association
```{r fig.width=10, fig.height=5}
pval_lst <- list()
for (i in 1:length(gibbs_lst)){
  perm_PM <- gibbs_lst[[i]]
  gibbs_res_tb <- make_gibbs_res_tb(perm_PM, all_KOs.G_mat, compute_pve = F)
  pval_matrix <- gibbs_res_tb %>% select(starts_with("pval"))
  pval_lst[[i]] <- unlist(pval_matrix)
}
summ_pvalues(unlist(pval_lst),
             title_text = paste0(length(gibbs_lst), " Permutations, 20 factors x 15 KOs"))
```

## LFSR

```{r lfsr function}
print_signif_tb <- function(signif_num, caption_text){
  if (length(signif_num) < 16){
    signif_num <- c(signif_num, rep(NA, 16 - length(signif_num)))
  }
  signif_num <- t(data.frame(KO = names(signif_num),
                             Num_genes = signif_num,
                             row.names = NULL))
  print(knitr::kable(rbind(signif_num[, 1:8], signif_num[, 9:16]),
                     caption = caption_text) %>%
    kable_styling() %>% scroll_box(width = '100%'))
}
print_lfsr_signif_tb <- function(lfsr_mat, lfsr_cutoff, caption_text){
  signif_num <- colSums(lfsr_mat < lfsr_cutoff)
  print_signif_tb(signif_num, caption_text)
}
```

```{r load LFSR results}
for (i in 1:length(gibbs_lst)){
  lfsr_mat <- readRDS(paste0(wkdir, gsfa_folder, "All.gibbs_obj_k20.perm_", i, ".lfsr_mat.rds"))
  print_lfsr_signif_tb(lfsr_mat, 0.05,
                       paste0("Permutation ", i, ", # of genes that passed LFSR cutoff of 0.05:"))
}
```
