---
title: "Local False Sign Rate Estimate of KO Effect on Genes"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/gene_discovery_functions.R")
print_signif_tb <- function(signif_num, caption_text){
  if (length(signif_num) < 30){
    signif_num <- c(signif_num, rep(NA, 30 - length(signif_num)))
  }
  signif_num <- t(data.frame(KO = names(signif_num),
                             Num_genes = signif_num,
                             row.names = NULL))
  print(knitr::kable(rbind(signif_num[, 1:10], signif_num[, 11:20], signif_num[, 21:30]),
               caption = caption_text) %>%
    kable_styling() %>% scroll_box(width = '100%'))
}
print_lfsr_signif_tb <- function(lfsr_mat, lfsr_cutoff, caption_text){
  signif_num <- colSums(lfsr_mat < lfsr_cutoff)
  print_signif_tb(signif_num, caption_text)
}

wkdir <- "/project2/xinhe/yifan/Factor_analysis/GSE108699_MCF10A/"
dge_cutoff <- 0.05
lfsr_cutoff <- 0.05
```

Both GSFA and DE analysis were conducted on genes filtered by the criterion of detection rate > 10% (8046 genes kept).

# Differential Gene Expression Test

We can compare the gene expression profiles of cells under each KO (gRNA) condition with those of cells without this gRNA using either **Wilcoxon rank sum test** or **Welch's t-test**.

We have 2 variations on the gene expression data we use: (1) **log-CPM expression**; (2) **corrected expression**: log-CPM data corrected for batch effect and other confounding factors and then scaled. The 2nd kind is the expression data we passed to GSFA.

## Wilcoxon rank sum test on log-CPM expression

The number of genes that passed an FDR threshold of 0.05 (Benjamini-Hochberg) under each KO condition is:

```{r wilcox_DE}
wilcox.DE_list <-
  readRDS(paste0(wkdir, "preprocess_output/DE_results.all_markers.seurat_wilcox.rds"))
wilcox.DE_list <- lapply(wilcox.DE_list, function(x){
    x <- x %>% mutate(gene_ID = rownames(x))
    x$fdr <- p.adjust(x$p_val, method = "fdr")
    return(x)
  })
DE_signif_counts <- sapply(wilcox.DE_list, function(x){filter(x, fdr < dge_cutoff) %>% nrow()})
print_signif_tb(DE_signif_counts,
                caption_text = paste0("Wilcoxon DGE on log(CPM) expression, FDR cutoff: ", dge_cutoff))
```

```{r wilcox_pval_FC_plot, fig.width=8, fig.height=14}
signif_markers <- names(DE_signif_counts)[DE_signif_counts > 0]
gridplot_dge_volcano(wilcox.DE_list[signif_markers],
                     dge_cutoff, title_text = "Wilcoxon DGE on log(CPM)")
```

## Welch's t-test on corrected expression

The number of genes that passed an FDR threshold of 0.05 (Benjamini-Hochberg) under each KO condition is:

```{r ttest_DE_corrected}
ttest_corr.DE_list <-
  readRDS(paste0(wkdir, "preprocess_output/DE_results.custom_ttest.detect_01.corrected_lib_NTC_singleton.rds"))
DE_signif_counts <-
  sapply(ttest_corr.DE_list, function(x){filter(x, fdr < dge_cutoff) %>% nrow()})
print_signif_tb(DE_signif_counts,
                caption_text = paste0("t-test DGE on corrected expression, FDR cutoff: ", dge_cutoff))
```

```{r ttest_pval_FC_plot, fig.width=12, fig.height=14}
ttest.merged_list <- merge_lists(wilcox.DE_list, ttest_corr.DE_list)
ttest.merged_list <- lapply(ttest.merged_list,
                            function(x){ x %>% rename(p_val = p_val.y, fdr = fdr.y) })
gridplot_dge_volcano(ttest.merged_list, dge_cutoff, title_text = "t-test DGE on corrected expression")
```

# LFSR Estimation from GSFA (NTC condition regressed out, singletons)

In terms of our GSFA method, we can also determine if a KO condition has an significant effect on gene expression in the following way:

For a given GSFA inference result, we estimated the local false sign rate (lfsr) of a certain knock-out effect on each gene based on the posteriors of $\beta$ and $W$.

For gene $j$ and KO condition $m$,
$$lfsr_{mj} = \text{min} \Big\{\text{Pr}(\sum_{k=1}^K \beta_{mk}W_{jk} \geq 0 \text{ | Data}), \text{Pr}(\sum_{k=1}^K \beta_{mk}W_{jk} \leq 0 \text{ | Data}) \Big\}$$
Under each KO condition, the number of genes that pass a certain LFSR cutoff would be compared to that from a the DGE tests mentioned above.

```{r no_NTC_lfsr}
no_NTC.lfsr_mat <- readRDS(paste0(wkdir, "gsfa_output_detect_01/sparse_beta_singleton/",
                           "lfsr/All.gibbs_obj_k20.svd.lfsr_mat.rds"))
print_lfsr_signif_tb(no_NTC.lfsr_mat, lfsr_cutoff,
                     paste0("GSFA NTC regressed, LFSR cutoff: ", lfsr_cutoff))
```

## Comparison between NTC-regressed and Wilcoxon DGE (log-CPM) results

```{r}
olap_df <- DGE_overlap_num(no_NTC.lfsr_mat, wilcox.DE_list,
                           lfsr_cutoff = lfsr_cutoff, dge_cutoff = dge_cutoff,
                           lfsr_name = "GSFA NTC-regressed",
                           dge_name = "Wilcoxon DGE")
```

## Comparison between NTC-regressed and DGE t-test (corrected) results

```{r}
olap_df <- DGE_overlap_num(no_NTC.lfsr_mat, ttest_corr.DE_list,
                           lfsr_cutoff = lfsr_cutoff, dge_cutoff = dge_cutoff,
                           lfsr_name = "GSFA NTC-regressed",
                           dge_name = "DGE t-test")
```

## GSFA estimated effect size vs LFSR

```{r load GSFA NTC-regressed inference data}
gibbs_PM <- readRDS(paste0(wkdir, "gsfa_output_detect_01/sparse_beta_singleton/",
                           "All.gibbs_obj_k20.svd.restart.PM.rds"))
no_NTC.DE_list <- compute_beta_dot_W(no_NTC.lfsr_mat, gibbs_PM)
signif_markers <- colnames(no_NTC.lfsr_mat)[colSums(no_NTC.lfsr_mat < 0.05) > 0]

Y <- readRDS(paste0(wkdir, "data/cropseq_dox.scaled_detect_01.corrected_lib_NTC_singleton.rds"))
Y <- t(Y)
dox_metadata <- readRDS(paste0(wkdir, "preprocess_output/dox_metadata.rds"))
dox_metadata <- dox_metadata %>% filter(guide_count > 0)
dox_metadata$type_count <- rowSums(dox_metadata[, 17:46])
dox_metadata <- dox_metadata %>% filter(type_count == 1)
G_mat <- dox_metadata[, 17:46] * 1
G_mat <- G_mat[, colnames(G_mat) != "NONTARGETING"]
```

### Compared to DGE on input Y

```{r fig.width=8, fig.height=8}
GSFA_Y.DE_list <- merge_lists(no_NTC.DE_list[signif_markers], ttest_corr.DE_list)
gridplot_betaW_lfsr(GSFA_Y.DE_list, title_text = "GSFA NTC-included")
```

To understand why t-test on the input matrix, $Y$, for GSFA still found much fewer genes than LFSR from factor analysis, we consider the factors that are significantly associated with each guide. We have 2 possible assumptions:

(1) The top factors captured by GSFA (i.e. $ZW^T$) might increase gene association with guide.

(2) The residuals of $Y$ after regressing out non-relevant factors (i.e. $\text{resid} (Y \sim Z_{l \in \{\hat\beta_l = 0\}})$) might increase gene association with guide.

### DGE p values on $ZW^T$


```{r ttest on ZWT}
denoised.DE_list <- list()
denoised_Y <- t(gibbs_PM$Z_pm %*% t(gibbs_PM$W_pm))
for (m in signif_markers){
  denoised.DE_df <- DE_test(expression_mat = denoised_Y,
                            condition = G_mat[, m],
                            gene_names = colnames(Y))
  denoised.DE_list[[m]] <- denoised.DE_df
}
```

```{r fig.width=8, fig.height=8}
GSFA_denoised_Y.DE_list <- merge_lists(no_NTC.DE_list, denoised.DE_list)
gridplot_betaW_lfsr(GSFA_denoised_Y.DE_list, title_text = "GSFA NTC-included")
```

### DGE p values on residuals of Y after associated modules were regressed out

```{r ttest on modules}
# signif.DE_list <- list()
resid.DE_list <- list()
for (m in signif_markers){
  marker_indx <- which(colnames(G_mat) == m)
  factor_indx <- which(gibbs_PM$Gamma_pm[marker_indx, ] > 0.8)
  if (length(factor_indx) > 0){
    cat("Knock-down gene:", m, "Associated factors:", paste(factor_indx, collapse = ","))
    cat("\n\n")
    # signif_modules_Y <- gibbs_PM$Z_pm[, factor_indx] %*% t(gibbs_PM$W_pm[, factor_indx])
    # signif.DE_df <- DE_test(expression_mat = t(signif_modules_Y),
    #                         condition = G_mat[, m],
    #                         gene_names = colnames(Y))
    # signif.DE_list[[m]] <- signif.DE_df
    resid_Y <- lm(Y ~ gibbs_PM$Z_pm[, -factor_indx])$residuals
    resid.DE_df <- DE_test(expression_mat = t(resid_Y),
                           condition = G_mat[, m],
                           gene_names = colnames(Y))
    resid.DE_list[[m]] <- resid.DE_df
  }
}
```

```{r fig.height=8, fig.width=8, eval=FALSE, include=FALSE}
GSFA_signif_modules.DE_list <- merge_lists(no_NTC.DE_list, signif.DE_list)
gridplot_betaW_lfsr(GSFA_signif_modules.DE_list, title_text = "GSFA NTC-regressed")
```

```{r fig.width=8, fig.height=8}
GSFA_resid_Y.DE_list <- merge_lists(no_NTC.DE_list, resid.DE_list)
gridplot_betaW_lfsr(GSFA_resid_Y.DE_list, title_text = "GSFA NTC-regressed")
```
