---
title: "Evaluation of GSFA on Simulated Data"
subtitle: "-- Realistic Settings"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 16, hjust = 0.5),
                             axis.title = element_text(size = 16),
                             axis.text = element_text(size = 14),
                             legend.title = element_text(size = 15),
                             legend.text = element_text(size = 14),
                             panel.grid.minor = element_blank())
)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/simulations/simulation_analytics.R")
wkdir <- paste0("/project2/xinhe/yifan/Factor_analysis/simulations/",
                "simulation_results/multi_factors_n_markers/realistic_beta_0.1_to_0.6/")
folders <- c("prior_sb_5.prior_rb_0.2.pi_0.05/",
             "prior_sb_5.prior_rb_0.2.pi_0.1/",
             "prior_sb_5.prior_rb_0.2.pi_0.2/")
settings <- c("pi=0.05", "pi=0.1", "pi=0.2")
marker_levels <- seq(0.1, 0.6, 0.1)
```

# Simulation Settings

We simulated our data in a bottom-up fashion:
$$G_{im} \overset{i.i.d.}{\sim} \text{Bern}(0.1),\hspace{3mm} \Phi_{ik} \overset{i.i.d.}{\sim} N(0, 1) \Rightarrow Z = G \beta + \Phi$$
$$ F_{jk} \sim \text{Bern}(\pi_j),\hspace{3mm}  U_{jk} \sim N(0, \sigma_w^2) \Rightarrow W = F \odot U$$
$$E_{ij} \sim N(0,\psi_j), Z, W \Rightarrow Y = ZW^T+E$$
$G \in \mathbb{R}^{N \times M}, \beta \in \mathbb{R}^{M \times K}, Z \in \mathbb{R}^{N \times K}, W \in \mathbb{R}^{P \times K}, Y \in \mathbb{R}^{N \times P}.$

For simulation cases in this report,

Sample size $N = 4000$, gene number $P = 6000$, factor number $K = 10$, and guide/marker number $M = 6$;    
$\sigma_w^2 = 0.5$, matrix $\beta$ takes the following form:

\begin{pmatrix}
0.1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0.2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0.3 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0.4 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0.5 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0.6 & 0 & 0 & 0 & 0 \\
\end{pmatrix}

The first 6 factors are each associated with a guide, with the effect sizes varying from 0.1 to 0.6, while the last 4 factors are not associated with any guides.

**We explored 3 cases when the density parameter $\pi_j$ is the same across $j$ but takes a value from 0.05, 0.1 to 0.2.**

Under each scenario, 500 random datasets were simulated, and both guided and unguided GSFA models were performed on each dataset for 2000 iterations starting from SVD initialization; posterior means were averaged over the last 1000 iterations. 

# Estimation of Effect Sizes ($\beta$s)

```{r load_beta_reg_and_rmse_ZZT}
paired_beta_plots <- list()
zzt_diff_list <- list()
for (i in 1:length(folders)){
  betaZ_list <- readRDS(paste0(wkdir, folders[i],
                               "regression_beta_n_rmse_ZZT.nrep_500.rds"))
  sub_text <- paste0("True Factor Density \u03C0=", betaZ_list$params$pi_true[1])
  paired_beta_plots[[i]] <- plot_paired_beta_reg_boxplots(betaZ_list$regression_stats,
                                                          marker_levels, sub_text) +
    scale_y_continuous(breaks = c(0.2, 0.4, 0.6, 0.8, 1),
                       minor_breaks = seq(0.1, 1, 0.1)) +
    theme(panel.grid.minor = element_line(size = 0.5))
  
  betaZ_list$rmse_ZZT$Setting <- settings[i]
  zzt_diff_list[[i]] <- betaZ_list$rmse_ZZT
}
```

```{r gridplot_beta_reg, fig.width=11, fig.height=8}
do.call(grid.arrange, c(paired_beta_plots, nrow = 2))
```

# Estimation of Factors ($Z$s)

We would like to evaluate how different our estimation of the factor matrix, $\hat{Z}$, is from the true value $Z$.   
Since the order of latent factors is non-identifiable, we focus on $ZZ^T$ and evaluate its estimation error using  $||\hat{Z}\hat{Z}^T - ZZ^T||_F/N$, where $||\cdot||_F$ is the Frobenius norm, and $N$ is the number of rows (samples) in $Z$.

```{r plot_rmse_ZZT, fig.height=5, fig.width=7}
plot_paired_rmse_zzt_boxplots(zzt_diff_list,
                              sub_text = paste0("various \u03C0s"))
```

## Correlation between matching $Z$ and $\hat{Z}$

On the other hand, since we know the true $Z$'s, we can match each of other inferred factors $\hat{Z}$'s 
with the true $Z$'s by maximum absolute Pearson correlation.

```{r load_matching_ZZ_corr}
Z_corr_across_pi <- list()
for (i in 1:length(folders)){
  Z_corr_across_pi[[i]] <- readRDS(paste0(wkdir, folders[i], "matched_ZZ_correlation.nrep_500.rds"))
}
```

```{r fig.height=4, fig.width=5, eval=FALSE, include=FALSE}
Z_corr_across_pi_df <- do.call(rbind, Z_corr_across_pi)
Z_corr_across_pi_df <- Z_corr_across_pi_df %>%
  mutate(pi = factor(pi)) %>% filter(Model == "Guided")
levels(Z_corr_across_pi_df$pi) <- settings
ggplot(Z_corr_across_pi_df, aes(x = pi, y = abs(Correlation), color = pi)) +
  geom_boxplot() +
  scale_color_brewer(palette = "Set2") +
  labs(x = "True Density of Factors",
       y = "|Correlation|",
       title = paste0("Correlation between matched true and inferred factors\n",
                      "(pooled over all factors)")) +
  theme(legend.position = "none")
```

```{r fig.height=5, fig.width=7}
Z_corr_across_pi_df <- do.call(rbind, Z_corr_across_pi)
Z_corr_across_pi_df <- Z_corr_across_pi_df %>%
  mutate(pi = factor(pi))
levels(Z_corr_across_pi_df$pi) <- settings
ggplot(Z_corr_across_pi_df, aes(x = pi, y = abs(Correlation), color = Model)) +
  geom_boxplot() +
  # scale_color_brewer(palette = "Set2") +
  scale_color_manual(values = c('#F8766D','#7CAE00')) +
  labs(x = "True Density of Factors",
       y = "|Correlation|",
       title = paste0("Correlation between matched true and inferred factors\n",
                      "(pooled over all factors)"))
```

```{r fig.height=10, fig.width=12}
ggplot(Z_corr_across_pi_df, aes(x = Factor, y = abs(Correlation), color = Model)) +
  geom_boxplot() +
  facet_wrap(~ pi, nrow = 2) +
  # scale_color_brewer(palette = "Set2") +
  scale_color_manual(values = c('#F8766D','#7CAE00')) +
  labs(y = "|Correlation|"
       # title = paste0("Correlation between matched true and inferred factors\n",
       #                "(\u03C0=", res_list$params$pi_true[1], ")")
       ) +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 16))
```

# Beta PIP Calibration

```{r fig.height=8, fig.width=12, eval=FALSE, include=FALSE}
# Beta PIP break down by marker
i <- 3
beta_pip_list <- readRDS(paste0(wkdir, folders[3], "beta_pip_list.nrep_500.rds"))
plot_beta_PIP_by_effect_size(beta_pip_list, M = 6,
                             title_text = paste0("Beta PIP Calibration, \u03C0=0.2"))
```

```{r fig.width=9, fig.height=8}
PIP_pooled_plots <- list()
for (i in 1:length(folders)){
  beta_pip_list <- readRDS(paste0(wkdir, folders[i], "beta_pip_list.nrep_500.rds"))
  PIP_pooled_plots[[i]] <- plot_beta_PIP_pooled(beta_pip_list, M = 6,
                                                title_text = paste0("Beta PIP Calibration (Pooled)\n",
                                                                    settings[i]))
}
do.call(grid.arrange, c(PIP_pooled_plots, nrow = 2))
```

# Gene Detection

## Detection methods

1.  GSFA + Local False Sign Rate (LFSR)

In terms of our GSFA method, we can estimate the effect of a guide/marker on each gene by summarizing over all the factors through the local false sign rate (LFSR):

For gene $j$ and guide $m$, based on the posteriors of $\beta$ and $W$ of an inference,
$$\text{LFSR}_{mj} = \text{min} \Big\{\text{Pr}(\sum_{k=1}^K \beta_{mk}W_{jk} \geq 0 \text{ | Data}), \text{Pr}(\sum_{k=1}^K \beta_{mk}W_{jk} \leq 0 \text{ | Data}) \Big\}$$

2.  Welch's t-test + FDR

Alternatively, we can simply test each gene directly, measuring whether there is an difference in $Y_j$ between the two groups of samples under guide $m$ using Welch's t-test.

## ROC

To obtain the following ROC curves, we varied the cutoff from 0 to 1 on both LFSR and t-test FDR values.

The actual cutoff values used are (correspond to the datapoints on an ROC curve from left to right)

```{r}
PR_list <- readRDS(paste0(wkdir, folders[1], "lfsr_vs_DE.PR_curves.nrep_500.rds"))
cutoffs <- paste0(unique(PR_list$lfsr_means$cutoff), collapse = ", ")
cat(cutoffs)
cat("\n")
```

### $\pi$ = 0.05

```{r fig.height=5, fig.width=7}
i <- 1
PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))

plot_paired_ROC(PR_list$lfsr_means, PR_list$ttest_means, "GSFA", "t-test",
                marker_levels = marker_levels) +
  labs(title = paste0("ROC -- GSFA LFSR vs t-test FDR thresholding\n",
                      "True Factor Density \u03C0=", PR_list$params$pi_true[1]))
```

### $\pi$ = 0.1

```{r fig.height=5, fig.width=7}
i <- 2
PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))

plot_paired_ROC(PR_list$lfsr_means, PR_list$ttest_means, "GSFA", "t-test",
                marker_levels = marker_levels) +
  labs(title = paste0("ROC -- GSFA LFSR vs t-test FDR thresholding\n",
                      "True Factor Density \u03C0=", PR_list$params$pi_true[1]))
```

### $\pi$ = 0.2

```{r fig.height=5, fig.width=7}
i <- 3
PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))

plot_paired_ROC(PR_list$lfsr_means, PR_list$ttest_means, "GSFA", "t-test",
                marker_levels = marker_levels) +
  labs(title = paste0("ROC -- GSFA LFSR vs t-test FDR thresholding\n",
                      "True Factor Density \u03C0=", PR_list$params$pi_true[1]))
```

```{r fig.width=12, eval=FALSE, include=FALSE}
# Precision-Recall
sub_text <- paste0("True Factor Density \u03C0=", PR_list$params$pi_true[1])
plot_paired_PR_boxplots(PR_list$lfsr_list$cutoff_0.05, PR_list$ttest_list$cutoff_0.05,
                        "GSFA LFSR<0.05", "t-test FDR<0.05",
                        marker_levels, sub_text)
```

## FDR Calibration at LFSR 0.05

### Pooled

```{r fig.width=5, fig.height=4}
pooled_fdr_list <- list()
for (i in 1:length(folders)){
  PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))
  precision_samples_df <- PR_list$lfsr_list$cutoff_0.05 %>% select(starts_with("precision"))
  pooled_fdr_list[[i]] <- data.frame(pooled_fdr = na.omit(unlist(1 - precision_samples_df)),
                                     setting = settings[i])
}
pooled_fdr_df <- do.call(rbind, pooled_fdr_list)

ggplot(pooled_fdr_df, aes(x = setting, y = pooled_fdr, color = setting)) +
  geom_boxplot() +
  geom_hline(yintercept = 0.05, color = "black", linetype = "dotted") +
  scale_color_brewer(palette = "Set2") +
  labs(x = "True Density of Factors",
       y = "Observed FDR of Genes",
       title = paste0("GSFA observed DEG FDR at LFSR cutoff of 0.05\n",
                      "(pooled over all guides, black line: 0.05)")) +
  theme(legend.position = "none")
```

### $\pi$ = 0.05

```{r}
i <- 1
PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))
plot_FDR_fixed_cutoff(PR_list$lfsr_list$cutoff_0.05, cutoff = 0.05,
                      marker_levels = marker_levels,
                      sub_text = paste0("(\u03C0 = ", PR_list$params$pi_true[1],
                                        ", red line: expected)"))
```

```{r fig.width=10, fig.height=6.5}
gridplot_lfsr_vs_observed_fdr(means_df = PR_list$lfsr_means,
                              sds_df = PR_list$lfsr_sds,
                              marker_levels = marker_levels,
                              title_text = paste0("LFSR Calibration (\u03C0 = ",
                                                  PR_list$params$pi_true[1],
                                                  ", mean±sd)"),
                              n_row = 2)
```

### $\pi$ = 0.1

```{r}
i <- 2
PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))
plot_FDR_fixed_cutoff(PR_list$lfsr_list$cutoff_0.05, cutoff = 0.05,
                      marker_levels = marker_levels,
                      sub_text = paste0("(\u03C0 = ", PR_list$params$pi_true[1],
                                        ", red line: expected)"))
```

```{r fig.width=10, fig.height=6.5}
gridplot_lfsr_vs_observed_fdr(means_df = PR_list$lfsr_means,
                              sds_df = PR_list$lfsr_sds,
                              marker_levels = marker_levels,
                              title_text = paste0("LFSR Calibration (\u03C0 = ",
                                                  PR_list$params$pi_true[1],
                                                  ", mean±sd)"),
                              n_row = 2)
```

### $\pi$ = 0.2

```{r}
i <- 3
PR_list <- readRDS(paste0(wkdir, folders[i], "lfsr_vs_DE.PR_curves.nrep_500.rds"))
plot_FDR_fixed_cutoff(PR_list$lfsr_list$cutoff_0.05, cutoff = 0.05,
                      marker_levels = marker_levels,
                      sub_text = paste0("(\u03C0 = ", PR_list$params$pi_true[1],
                                        ", red line: expected)"))
```

```{r fig.width=10, fig.height=6.5}
gridplot_lfsr_vs_observed_fdr(means_df = PR_list$lfsr_means,
                              sds_df = PR_list$lfsr_sds,
                              marker_levels = marker_levels,
                              title_text = paste0("LFSR Calibration (\u03C0 = ",
                                                  PR_list$params$pi_true[1],
                                                  ", mean±sd)"),
                              n_row = 2)
```

