---
title: "Guided Factor Analysis on Unstimulated T Cell CROP-seq Data"
subtitle: "-- Deviance residual transformed + normal-mixture prior"
author: "Yifan Zhou (zhouyf@uchicago.edu)"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
requireNamespace("pander", quietly = TRUE)
library(data.table)
library(Matrix)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(plot.title = element_text(size = 14, hjust = 0.5),
                             axis.title = element_text(size = 14),
                             axis.text = element_text(size = 12),
                             legend.title = element_text(size = 13),
                             legend.text = element_text(size = 12),
                             panel.grid.minor = element_blank())
)
library(ggrepel)
library(gridExtra)
library(ComplexHeatmap)
library(kableExtra)

# set default chunk output
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA,
                      tidy = FALSE,
                      fig.width = 8,
                      fig.height = 6,
                      fig.align = "center",
                      results = "asis")

# formatting of pander tables
pander::panderOptions('knitr.auto.asis', FALSE)
pander::panderOptions("table.split.table", Inf)
```

```{r}
source("/project2/xinhe/yifan/Factor_analysis/reports/R/analytics_functions.R")
print_signif_tb <- function(signif_num, caption_text){
  signif_num <- t(data.frame(KO = names(signif_num),
                             Num_genes = signif_num,
                             row.names = NULL))
  knitr::kable(rbind(signif_num[, 1:7], signif_num[, 8:14], signif_num[, 15:21]),
                     caption = caption_text) %>%
    kable_styling() %>% scroll_box(width = '100%')
}
wkdir <- "/project2/xinhe/yifan/Factor_analysis/Stimulated_T_Cells/"
gsfa_folder <- "gsfa_output_detect_01/unstim_normal_mixture/"
guide <- "All"
pip_col_fun <- circlize::colorRamp2(breaks = c(0, 0.5, 1),
                                    colors = c("black", "purple", "gold"))
```

# Single cell expression data

Source:   
[Genome-wide CRISPR Screens in Primary Human T Cells Reveal Key Regulators of Immune Function](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6689405/), GEO accession: GSE119450.

Perturbations:    
CRISPR knock-out of 20 genes (2 gRNAs per gene) + 8 non-targeting gRNAs. Guide conditions were defined on the target gene level; target genes were either found to regulate T cell responses in the genome-wide screens, or known checkpoint genes.    
Guide RNAs were introduced into T cells through a novel procedure called sgRNA lentiviral infection with Cas9 protein electroporation (SLICE).

Cells:    
Primary human CD8+ T cells from two healthy donors, **without** T cell receptor (TCR) stimulation.   
Cells from 2 donors were pooled together into 1 analysis. All cells have only a single type of gRNA readout. Quality control resulted in 10644 cells.

Genes:    
**Top 6000 genes ranked by the multinomial deviance statistics were kept.**

Normalization:    
Deviance residual transformation.   
**Batch effect, unique UMI count, library size, and mitochondria percentage were all corrected for. The corrected and scaled expression data were used as input for subsequent factor analysis.**

```{r load meta data}
metadata <- readRDS(paste0(wkdir, "processed_data/metadata.unstimulated_merged.rds"))
all_KOs.G_mat <- metadata[, 4:24]
KO_names <- colnames(all_KOs.G_mat)
```

# Guided results

Here, our "guide", $G$ matrix, consists of 21 types (20 genes + negative control) of KO conditions across cells.

```{r}
if (guide == "no_NTC"){
  KO_names <- KO_names[KO_names != "NonTarget"]
  all_KOs.G_mat <- all_KOs.G_mat[, KO_names]
}
nmarkers <- ncol(all_KOs.G_mat)
```

In each case, Gibbs sampling was conducted for 3000 iterations, and the posterior mean estimates were averaged over the last 1000 iterations.

## SVD Initialization
```{r svd_tb}
All_KOs.gibbs_PM <- readRDS(paste0(wkdir, gsfa_folder, guide, ".gibbs_obj_k20.svd.restart.PM.rds"))
stopifnot(nrow(All_KOs.gibbs_PM$beta_pm) == nmarkers + 1)
All_KOs.gibbs_res_tb <- make_gibbs_res_tb(All_KOs.gibbs_PM, all_KOs.G_mat, compute_pve = F)
```

### Local False Sign Rate (LFSR)

For a given GSFA inference result, we can estimate the effect a certain KO condition $m$ has on the expression of gene $j$ by computing the LFSR of $\beta_{m\cdot} \cdot W_{j\cdot}$.

Number of genes that passed GSFA LFSR < 0.05 under each perturbation:
```{r}
lfsr_cutoff <- 0.05
lfsr_mat <- readRDS(paste0(wkdir, gsfa_folder, guide, ".gibbs_obj_k20.svd.restart.lfsr_mat.rds"))
lfsr_signif_num <- colSums(lfsr_mat < lfsr_cutoff)
print_signif_tb(lfsr_signif_num, paste0("GSFA, # of genes that passed LFSR cutoff of 0.05:"))
```

Number of genes that passed t-test FDR < 0.05 under each perturbation:
```{r ttest_DE}
fdr_cutoff <- 0.05
DE_list <- readRDS(paste0(wkdir, "processed_data/DE_results.custom_ttest.unstimulated_merged.detect_01.corrected.rds"))
DE_signif_counts <- sapply(DE_list, function(x){filter(x, fdr < fdr_cutoff) %>% nrow()})
print_signif_tb(DE_signif_counts, paste0("Welch's t-test, # of genes that passed FDR cutoff of 0.05:"))
```

### Estimate of Factor ~ Perturbation Associations ($\beta$)

```{r fig.width=9, fig.height=7}
reorder_KO <- c("ARID1A", "CBLB", "CD5", "RASA2", "SOCS1", "TCEB2", "CD3D", "LCP2",
                "BTLA", "C10orf54", "CDKN1B", "DGKA", "DGKZ", "HAVCR2", "LAG3", "MEF2D",
                "PDCD1", "STAT6", "TMEM222", "TNFRSF9", "NonTarget")

dotplot_beta_PIP(t(All_KOs.gibbs_PM$Gamma_pm), t(All_KOs.gibbs_PM$beta_pm),
                 marker_names = colnames(all_KOs.G_mat), reorder_markers = reorder_KO)
```

### Factor ~ KO Posterior Association
```{r fig.width=10}
heatmap_matrix <- All_KOs.gibbs_res_tb %>% select(starts_with("pval"))
rownames(heatmap_matrix) <- 1:nrow(heatmap_matrix)
colnames(heatmap_matrix) <- colnames(all_KOs.G_mat)
plot_pval_heatmap(heatmap_matrix,
                  factor_annot = All_KOs.gibbs_res_tb$pi)
guided.heatmap_matrix <- heatmap_matrix
gsfa_KO_names <- apply(guided.heatmap_matrix, 1,
                       function(x){ paste(colnames(all_KOs.G_mat)[x < 1e-4], collapse = ",") })
gsfa_KO_names <- paste(1:nrow(guided.heatmap_matrix), gsfa_KO_names, sep = " - ")
```

```{r fig.width=10, fig.height=5}
summ_pvalues(unlist(heatmap_matrix),
             title_text = "GSFA All KOs (K=20, SVD)")
```

### Beta PIP vs P-Value
```{r fig.width=12, fig.height=5}
beta_pip_matrix <- t(All_KOs.gibbs_PM$Gamma_pm[-nrow(All_KOs.gibbs_PM$Gamma_pm), ])
p1 <- qplot(unlist(beta_pip_matrix), -log10(unlist(heatmap_matrix)),
            xlab = "Beta PIP", ylab = "-log10(Association p-value)",
            main = "20 factors x 15 conditions")
p2 <- qplot(unlist(beta_pip_matrix), -log10(unlist(heatmap_matrix)),
            ylim = c(0, 10),
            xlab = "Beta PIP", ylab = "-log10(Association p-value)",
            main = "Y-axis truncated at 10")
grid.arrange(p1, p2, nrow = 1)
```

### Correlation within Factors

```{r fig.width=7, fig.height=5}
plot_pairwise.corr_heatmap(input_mat_1 = All_KOs.gibbs_PM$Z_pm,
                           name_1 = "Pairwise Correlation in Sample Loadings - Z (GSFA)",
                           corr_type = "pearson",
                           label_size = 10)
```

```{r fig.width=7, fig.height=5}
plot_pairwise.corr_heatmap(input_mat_1 = (All_KOs.gibbs_PM$F_pm > 0.95) * 1,
                           corr_type = "jaccard",
                           name_1 = "Binarized Gene Loadings - F (PIP > 0.95)")
```

### Correlation w/ Factors from Stimulated Result

```{r}
unstim_exp_file <-  "processed_data/deviance_residual.unstim_merged_top_6k.corrected_4.scaled.rds"
unstim_exp <- readRDS(paste0(wkdir, unstim_exp_file))
stim_exp_file <-  "processed_data/deviance_residual.stimulated_merged_top_6k.corrected_4.scaled.rds"
stim_exp <- readRDS(paste0(wkdir, stim_exp_file))
common_genes <- intersect(rownames(stim_exp), rownames(unstim_exp))
unstim_index <- match(common_genes, rownames(unstim_exp))
stim_index <- match(common_genes, rownames(stim_exp))
stim.gibbs_PM <- readRDS(paste0(wkdir, "gsfa_output_detect_01/stim_normal_mixture/",
                                guide, ".gibbs_obj_k20.svd.restart.PM.rds"))
```

```{r fig.width=7, fig.height=5}
plot_pairwise.corr_heatmap(input_mat_1 = All_KOs.gibbs_PM$W_pm[unstim_index, ],
                           input_mat_2 = stim.gibbs_PM$W_pm[stim_index, ],
                           name_1 = "Gene Loadings - W (GSFA Unstimulated)",
                           name_2 = "Gene Loadings - W (GSFA Stimulated)",
                           corr_type = "pearson",
                           label_size = 10)
```

# Gene Ontology Enrichment Analysis by Factor

Target genes: Genes w/ non-zero loadings in each factor (PIP cutoff at 0.95);    
Backgroud genes: all 6000 genes used in factor analysis;    
Statistical test: hypergeometric test (over-representation test);    
Only GO terms/pathways that satisfy fold change $\geq$ 2 and test FDR $<$ 0.05 are shown below.

```{r}
feature.names <- data.frame(fread(paste0(wkdir, "GSE119450_RAW/D1N/genes.tsv"),
                                  header = FALSE), stringsAsFactors = FALSE)
names(feature.names) <- c("ID", "Name")
```

## GO Slim Over-Representation Analysis
    
Gene sets: [Gene ontology](http://www.geneontology.org/) "Biological Process" (non-redundant).

```{r go_bp}
gsfa.Web_ORA_go_bp <-
  readRDS(paste0(wkdir, gsfa_folder,
                 "WebGestalt_ORA/All.gibbs_obj_k20.svd.go_bp.cutoff_0.95.rds"))
signif_num <- print_enrich_ORA_tb(gsfa.Web_ORA_go_bp, enrich_type = "GO terms",
                                  convert_genes = T, gene_map = feature.names)
signif_num.df <- t(data.frame(Factor = 1:length(signif_num),
                              Signif_GO_terms = signif_num,
                              row.names = NULL))
knitr::kable(signif_num.df, caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '100%')
```

### Factors of interest

Immune response:  
3, 6, 10, 11, 12, 13, 18, 19

Cell cycle:  
7, 18  

**From the reference study:**

Targeting CD5, RASA2, SOCS1, and CBLB promoted programs that are characterized by the induction of markers of activation states (e.g. IL2RA, TNFRSF18/GITR), cell cycle genes (e.g. MKI67, UBE2S, CENPF and TOP2A), and effector molecules (e.g GZMB).

**Associated factors in GSFA:**

RASA2: 10, 12, 16   
CDKN1B: 2

**From the reference study:**

Targeting CD3D or LCP2 inhibited the cluster 10 activation program and promoted programs characterized by expression of resting state markers such as IL7R and CCR7.

**Associated factors in GSFA**

## Reactome Pathway Over-Representation Analysis

Gene sets: The [Reactome pathway database](http://www.reactome.org/).

```{r reactome}
gsfa.Web_ORA_react <-
  readRDS(paste0(wkdir, gsfa_folder,
                 "WebGestalt_ORA/All.gibbs_obj_k20.svd.Reactome.cutoff_0.95.rds"))

signif_num <- print_enrich_ORA_tb(gsfa.Web_ORA_react, enrich_type = "Reactome pathways",
                                  convert_genes = T, gene_map = feature.names)
signif_num.df <- t(data.frame(Factor = 1:length(signif_num),
                              Signif_GO_terms = signif_num,
                              row.names = NULL))
knitr::kable(signif_num.df, caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '100%')
```

# Gene Set Enrichment Analysis by Perturbation

Target genes: Genes w/ LFSR < 0.05 under each perturbation marker gene;    
Backgroud genes: all 6000 genes used in factor analysis;    
Statistical test: hypergeometric test (over-representation test);    
Only GO terms/pathways that satisfy fold change $\geq$ 2 and test FDR $<$ 0.05 are shown below.

## GO Slim Over-Representation Analysis

Gene sets: [Gene ontology](http://www.geneontology.org/) "Biological Process" (non-redundant).

```{r go_bp_by_marker}
Web_ORA_go_bp.by_marker <-
  readRDS(paste0(wkdir, gsfa_folder,
                 "WebGestalt_ORA/All.gibbs_obj_k20.svd.go_bp.by_marker.lfsr_0.05.rds"))
signif_num <- print_enrich_ORA_tb(Web_ORA_go_bp.by_marker,
                                  enrich_type = "GO terms", list_type = "per_marker",
                                  convert_genes = T, gene_map = feature.names)
signif_num.df <- t(data.frame(Marker = names(Web_ORA_go_bp.by_marker),
                              Signif_GO_terms = signif_num,
                              row.names = NULL))
knitr::kable(signif_num.df, caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '100%')
```

## Reactome Pathway Over-Representation Analysis

Gene sets: The [Reactome pathway database](http://www.reactome.org/).

```{r reactome_by_marker}
Web_ORA_react.by_marker <-
  readRDS(paste0(wkdir, gsfa_folder,
                 "WebGestalt_ORA/All.gibbs_obj_k20.svd.Reactome.by_marker.lfsr_0.05.rds"))

signif_num <- print_enrich_ORA_tb(Web_ORA_react.by_marker,
                                  enrich_type = "Reactome pathways", list_type = "per_marker",
                                  convert_genes = T, gene_map = feature.names)
signif_num.df <- t(data.frame(Marker = names(Web_ORA_react.by_marker),
                              Signif_GO_terms = signif_num,
                              row.names = NULL))
knitr::kable(signif_num.df, caption = "Summary:") %>%
  kable_styling() %>% scroll_box(width = '100%')
```
